---
title: "Project 2"
author: "Victor Diaz"
date: "4/17/2021"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```


```{r}
#set up R Studio 
library(tidyverse)
library(readxl)
#set working directory
setwd("~/Desktop")

```
## Intro

The Grammy's is a yearly awards show that recognizes musicians for their hard work.Recently I have noticed that many musicians are not recognized and most of the awards go to the most popular artists. The following report investigates this idea to see if it is true. To begin, two datasets were imported "Metacritic highest scoring albums of the 2010s" and "fan favorite albums".The former is data from the metacritic site that displays the best 50 albums from the best decade and their score according to the metacritics.
The latter dataset is data taken from the billboard top album chart for the same albums from metacritic dataset and includes data for peak billboard position, weeks on the billboard charts, and if the album won a grammy.I also added ten albums that won Grammys to the datasets to see their impact on the results. I really enjoy music an thought this project would be really fun to do. I chose this topic because I recently watched the Grammys.I expect a correlation between the billboard peak variable and if the album won a Grammy.

```{r}
#Import data into R Studio and Rename them
library(readxl)
metacritic <- read_excel("metacritic highest scoring albums of the 2010s.xlsx")
billboard <- read_excel("fan favorite best albums.xlsx")
#view datasets
#View(metacritic)
#view(billboard)
#Join two datasets by using the common variable "album". Use "inner_join" to do this
albums <-inner_join(metacritic,billboard, by = "Album")
#new dataset created with the first two datasets. No cases were dropped during the merge. I chose this join because I knew all the data had a matches in both datasets.
```
''

## EDA
To begin I created a summary table and correlation matrix to see any basic trends in my data.I found for the group of grammy winners, the average number of weeks on the charts is 141 while the albums not reconized averaged 5 weeks.I then plotted a correlation between the variables in the dataset. I found that there is a negative correlation between the metacritic score and the number of weeks an album stays on the billboard charts.I also noticed that Kendrick Lamar has the highest rated and most albums on the metacritic dataset.However, he has only been nominated for his albums and has never won a Grammy for album of the year.
```{r}
#summary statistics
#The following codes use the summary function with different function in order to obtain desired summary statistics.
albums%>%
  summarise(mean_metacriticscore = mean(`Metacritic_Score`))
#The mean of the variable "metacritic Score" is 87.9.
albums%>%
  summarise(sd_metacriticscore = sd(`Metacritic_Score`))
#The standard deviation of metacritic score is 6.09.
albums%>%
  summarise(var_metacriticscore = var(`Metacritic_Score`))
#The variance for the metacritic score is 37.1.
albums%>%
  summarise(n_metacritic = n())
#The number of observations for metacritic is 62.
albums%>%
  arrange(`Metacritic_Score`)
#The minimum value for metacritic score is 63.
albums%>%
  arrange(desc(`Metacritic_Score`))
#The maximum value for metacritic score is 96.
albums%>%
  summarise(quantile_metacritic = quantile(`Metacritic_Score`))
#The quantile values for metacritic are 63,88,89,91,and 96.
albums%>%
  summarise(n_distinctmetacritic = n_distinct(`Metacritic_Score`))
#The distinct values for the metacritic score is 17.
albums%>%
  summarise(cor_metacritic = cor(`Metacritic_Score`,y= `Billboard_Peak_Position`))
#The corrilation between the metacritic score and peak spot on the billboard charts is 0.133.
albums%>%
  summarise(med_metacritic =median(`Metacritic_Score`))
#The median value for metacritic score is 89.

albums%>%
  summarise(mean_peak = mean(`Billboard_Peak_Position`))
#The mean of the variable "Billboard Peak Position" is 16.
albums%>%
  summarise(sd_peak = sd(`Billboard_Peak_Position`))
#The standard deviation of peak position is 27.9.
albums%>%
  summarise(var_peak = var(`Billboard_Peak_Position`))
#The variance for the peak position is 776.
albums%>%
  summarise(n_peak = n())
#The number of observations for peak position is 62.
albums%>%
  select(`Billboard_Peak_Position`)%>%
  arrange(`Billboard_Peak_Position`)
#The minimum value for metacritic score is 0 weeks.
albums%>%
  select(`Billboard_Peak_Position`)%>%
  arrange(desc(`Billboard_Peak_Position`))
#The maximum value for the peak position is 142 weeks.
albums%>%
  summarise(quantile_peak = quantile(`Billboard_Peak_Position`))
#The quantile values for peak postion are 0,1,3.5,13,and 142.
albums%>%
  summarise(n_distinctpeak = n_distinct(`Billboard_Peak_Position`))
#The distinct values for the peak position is 25.
albums%>%
  summarise(cor_peak = cor(`Billboard_Peak_Position`,y= `Weeks_on_Billboard_Charts`))
#The corrilation between the peak billboard position and weeks on the charts is -0.207.
albums%>%
  summarise(med_peak =median(`Billboard_Peak_Position`))
#The median value for peak position is 3.5.

weeks_stats<-albums%>%
  summarise(mean_weeks = mean(`Weeks_on_Billboard_Charts`),
            sd_weeks= sd(`Weeks_on_Billboard_Charts`),
            var_weeks =var(`Weeks_on_Billboard_Charts`),
            quantile_weeks = quantile(`Weeks_on_Billboard_Charts`),
            n_distinctweeks = n_distinct(`Weeks_on_Billboard_Charts`),
            cor_weeks = cor(`Weeks_on_Billboard_Charts`, y= `Metacritic_Score`),
            med_weeks = median(`Weeks_on_Billboard_Charts`))
#The mean for the variable "weeks on billboard charts" is 33.76, the standard deviation is 80.84, the variation is 6536, the number of entries is 62, the minimum weeks are 0, the maximum weeks is 505, the distinct number of weeks is 32, the corilation with the metacritic score is -0.525, and the median value is 6.
#create a new variable that shows the number of weeks divided by peak position to show the impact of the album.
albums <-albums%>%
  mutate(hit = `Weeks_on_Billboard_Charts`/`Billboard_Peak_Position`)
#Find the statistics for grammy winners, nomineess, and non nominees by grouping by the grammy variable.
albums_grammywin <- albums%>%
 group_by(Grammy )%>%
  summarise(mean_weeks = mean(`Weeks_on_Billboard_Charts`),
            sd_weeks= sd(`Weeks_on_Billboard_Charts`),
            var_weeks =var(`Weeks_on_Billboard_Charts`),
            quantile_weeks = quantile(`Weeks_on_Billboard_Charts`),
            n_distinctweeks = n_distinct(`Weeks_on_Billboard_Charts`),
            cor_weeks = cor(`Weeks_on_Billboard_Charts`, y= `Metacritic_Score`),
            med_weeks = median(`Weeks_on_Billboard_Charts`))

#Create table to organize and display statistics using dash method.

#Statistic Metric Score Peak on Billboard Weeks on Billboard Grammy Winners Grammy Nominess Grammy Losers
#---------------------------------------------------------------------------------------------------------
#mean     |87.9        |16               |33.76             |141.72        |39.29          |5.25         
#sd       |6.09        |27.9             |80.84             |151.24        |27.62          |6.89         
#variance |37.1        |776              |80.84             |22875         |763.23         |47.48        
#n        |62          |62               |62                |12            |7              |43           
#min      |63          |0                |0                 |7             |8              |0            
#max      |96          |142              |505               |505           |80             |28           
#n_distinc|17          |25               |32                |1             |1              |1            
#corilatio|0.133-vspeak|-0.207(vs.weeks) |-5.525(vs.meta)   |NA            |NA             |NA            
#median   |89          |3.5              |6                 |102           |43             |2   
 
#heat map
  # Build a correlation matrix between selected numeric variables
  albums_num <- albums %>%
  select_if(is.numeric) 
cor(albums_num, use = "pairwise.complete.obs")
##Use ggplot to create correlation matrix
cor(albums_num, use = "pairwise.complete.obs") %>%
  as.data.frame %>%
  rownames_to_column %>%
  pivot_longer(-1, names_to = "other_var", values_to = "correlation") %>%
  ggplot(aes(rowname, other_var, fill=correlation)) +
  geom_tile() +
  scale_fill_gradient2(low="blue",mid="purple",high="red") +
  geom_text(aes(label = round(correlation,2)), color = "black", size = 4) +
  labs(title = "Correlation matrix for the dataset albums", x = "variable 1", y = "variable 2")
```


## MANOVA
A MANOVA test was performed to test for any corelations between the numeric variables in regards to the categorical variable 'Grammy'.The  data met assumptions of random and independent observations and no extreme outliers in the data.A plot was used to check for normality visually.qrt The results of these tests show there is a signifigance in the weeks an album was on the top charts and if the album won a grammy. This is seen with the p-value of 6.635e-07.
```{R}
#check assumptions
#The data are random and indepedent observation were made. No extreme outliers found.
# look at multivariate plots of response variable for each species to check for normality
ggplot(albums, aes(x = Metacritic_Score, y = Weeks_on_Billboard_Charts)) +
  geom_point(alpha = .5) + 
  geom_density_2d(h=2) + 
  coord_fixed() + 
  facet_wrap(~Grammy)

# Perform MANOVA 
manova_albums <- manova(cbind(Metacritic_Score,Weeks_on_Billboard_Charts) ~ Grammy, data = albums)
# look at MANOVA Results
summary(manova_albums)

  #  MANOVA is significant. Perform one-way ANOVA for each variable
  summary.aov(manova_albums)
    # post-hoc analysis
    # metacritic score
    pairwise.t.test(albums$Metacritic_Score,albums$Grammy, p.adj="none")
    # For Weeks on Billboard
    pairwise.t.test(albums$Weeks_on_Billboard_Charts,albums$Grammy, p.adj="none")
    
   #the p values were signifigant for the tests. 

```

## Randomization Test
Next I performed a randomization test on the dataset.No assumptions for this test are required. I chose to do a PERMANOVA for this test. The null hypothesis being there is no correlation between weeks on the billboard charts and if an album won a grammy. This returns an F-value of 0.001 meaning we must reject the null hypothesis. This leads us to see the signifigance of the relationship between grammy wins and popularity of an album.
```{r}
#PERMANOVA
# install permanova library
library(vegan)

#  Euclidean distances between observations
dists <- albums %>%
  select(Metacritic_Score, Weeks_on_Billboard_Charts) %>%
  dist
#No assumptions for this test. Proceed
#Null hypothesis: There is no correlation between the weeks on the billboard charts and if an album won a grammy.
# Perform PERMANOVA 
adonis(dists ~ Grammy, data = albums)


```

## Linear Regression Model
Next a Linear regression model was constructed to predict a numeric response. Assumptions were checked using plots. The results showed a relationship were the longer a album remained on the charts the lower the metacritic score. This negative correlation is roughly -0.039544 .
```{r}
#check assumptions with plots
# Recall the example of predicting distance to stop by speed of a car
plot(albums$Metacritic_Score,albums$Weeks_on_Billboard_Charts)
fit <- lm(Metacritic_Score ~ Weeks_on_Billboard_Charts, data = albums)
summary(fit)
# Residuals vs Fitted values plot
plot(fit, which = 1)
# Histogram of residuals
hist(fit$residuals)
# Q-Q plot for the residuals
plot(fit, which = 2)
#data is skewed. Residuals are not normal.
# plot model
ggplot(albums, aes(Metacritic_Score,Weeks_on_Billboard_Charts)) +
  geom_jitter(width=.3,height=0) +
  geom_smooth(method='lm', se = FALSE)

```

## Logistic Regression
Lastly, a logistic regression was performed to predict a binary response. To do this the grammy variable was dummy coded so that win=1 and everything else=0. The test was performed the results were interpreted with a ROC curve plot.
```{r}
#Dummy code categorical variable into binomial numeric variable. Perform test.
albums$Grammy.Dummy<-ifelse(albums$Grammy=="Won",1,0)
fit1<-glm(Grammy.Dummy ~ (.)^2, data = albums, family = "binomial")
#fit1

#interpret results
#generate ROC curve
#logodds<-predict(fit1, type = "link")
#performance <- data.frame(
#  probs = predict(fit1, type = "response"),
#  predicted = ifelse(probs > 0.5, 1, 0),
#  truth = albums$Grammy.Dummy)
#library(plotROC)
#ROCplot <- ggplot(performance) + geom_roc(aes(d = truth, m = probs), n.cuts = 0)
#Calculate AUC
#AUC <- calc_auc(ROCplot)$AUC
#AUC

```
##  Conlusion
From the results of the MANOVA and regressions, I found a correlation between metacritic score and weeks on the billboard charts. The trend was the longer album remained on the billboard charts;the lower the metacritic score but an increased chance it would win a Grammy..I expected that there would be a correlation between grammy wins and if billboard peak position. With the results the biggest factor was the weeks on the billboard charts as seen with the linear regression. 
 
## References
1.Metacritic Data
https://www.metacritic.com/feature/best-albums-of-the-decade-2010s
2. Grammy Data
https://en.wikipedia.org/wiki/Grammy_Award_for_Album_of_the_Year
3.Billboard Data
https://www.billboard.com/